---
title: "Fitting Exercise"
---

# Analysis of the Mavoglurant pharmacokinetic data
## Loading data and initial exploration
```{r, warning=FALSE}
# Load packages
library(tidyverse)
library(tidymodels)
library(GGally)
library(here)
# Load data
data_raw <- read_csv(here("fitting-exercise", "Mavoglurant_A2121_nmpk.csv"))

# quick cheak
glimpse(data_raw)
```

```{r}
# Plot DV over TIME for each individual, colored by dose
ggplot(data_raw, aes(x = TIME, y = DV, group = ID, color = factor(DOSE))) +
  geom_line(alpha = 0.4) +
  labs(
    title = "Drug concentration over time by individual",
    x = "Time",
    y = "DV",
    color = "Dose"
  )
```

From this figures, we can find that the three dose levels show similar trends, with systemic exposure increasing relative to the dose.

```{r}
# keep one OOC = 1
data_OOC1 <- data_raw %>%
  filter(OCC == 1)

# Create a new data frame with the sum of DV for each individual, excluding time 0
y_dt <- data_OOC1 %>%
  filter(TIME != 0) %>%
  group_by(ID) %>%
  summarize(Y = sum(DV, na.rm = TRUE), .groups = "drop")

# keep time 0 data
time0 <- data_OOC1 %>%
  filter(TIME == 0) 

# Merge the two data frames to include the sum of DV and the time 0 data
d1 <- time0 %>%
  left_join(y_dt, by = "ID")

# Converts RACE and SEX to factor variables and keeps only these variables: Y,DOSE,AGE,SEX,RACE,WT,HT
d1 <- d1 %>%
  mutate(
    RACE = as.factor(RACE),
    SEX = as.factor(SEX),
    DOSE = as.factor(DOSE)
  ) %>% 
  select(Y, DOSE, AGE, SEX, RACE, WT, HT)

# Check the data
glimpse(d1)
head(d1)
summary(d1)
```

The dataset represents a study population of 198 participants with a notable demographic skew: 87% are Sex(group 1) and 62% belong to Race 1. While doses are centered around a mean of 37.4 mg, the majority of subjects received the low (25 mg) or high (50 mg) levels, leaving the medium dose less represented.

## Exploratory data analysis
### PLot
I next examined the relationship between the outcome Y and the predictors using scatterplots for continuous variables and boxplots for categorical variables.

```{r}
# Boxplot of Y by dose group
ggplot(d1, aes(x = factor(DOSE), y = Y)) +
  geom_boxplot() +
  labs(
    title = "Total drug concentration (Y) by dose group",
    x = "Dose group",
    y = "Y"
  )
```
A positive correlation is observed between dose levels and Outcome Y.
```{r}
# Scatterplot of Y versus AGE
ggplot(d1, aes(x = AGE, y = Y)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Total drug concentration (Y) versus age",
    x = "Age",
    y = "Y"
  )
```
There is no apparent correlation between Outcome Y and participant age.

```{r}
# Scatterplot of Y versus WT
ggplot(d1, aes(x = WT, y = Y)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Total drug concentration (Y) versus weight",
    x = "Weight",
    y = "Y"
  )
```
There is a slight negative correlation between Outcome Y and body weight.
```{r}
# Scatterplot of Y versus HT
ggplot(d1, aes(x = HT, y = Y)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Total drug concentration (Y) versus height",
    x = "Height",
    y = "Y"
  )
```
There is a slight negative correlation between Outcome Y and height.
```{r}
# Boxplot of Y by SEX
ggplot(d1, aes(x = SEX, y = Y)) +
  geom_boxplot() +
  labs(
    title = "Total drug concentration (Y) by sex",
    x = "Sex",
    y = "Y"
  )
```
There is no apparent difference between Outcome Y and participant age.

```{r}
# Boxplot of Y by RACE
ggplot(d1, aes(x = RACE, y = Y)) +
  geom_boxplot() +
  labs(
    title = "Total drug concentration (Y) by race",
    x = "Race",
    y = "Y"
  )
```
There is no apparent difference between Outcome Y and Race. 

### Distributions
```{r}
d1 %>%
  select(Y, AGE, WT, HT) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ variable, scales = "free") +
  labs(
    title = "Distributions of continuous variables",
    x = "Value",
    y = "Count"
  )
```
AGE: Right skewed, mostly concentrated between 20–35 years old.
HT: Roughly normally distributed, centered around 1.70–1.85 m.
WT: Slightly right skewed, most individuals weigh between 70–90 kg.
Y: Right-skewed with a long tail, most values fall between 1000–4000.
### Pair Plot
```{r}
d1 %>%
  select(Y, AGE, WT, HT) %>%
  ggpairs()
```
Based on the correlation matrix: WT and HT show the strongest positive correlation (r = 0.600), which is physiologically expected. Y has only weak correlations with all other variables, with WT being the most notable (r = -0.213). AGE and HT show a moderate negative correlation (r = -0.352).

## Model fitting
### Fit a linear model for Y using DOSE as the main predictor
```{r}
# Specify a linear regression model
lm_spec <- linear_reg() %>%
  set_engine("lm")

# Fit a linear model with DOSE as the predictor
lm_dose_fit <- lm_spec %>%
  fit(Y ~ DOSE, data = d1)

# View model results
lm_dose_fit
```
```{r}
# Generate predictions from the DOSE only linear model
lm_dose_results <- d1 %>%
  select(Y) %>%
  bind_cols(predict(lm_dose_fit, new_data = d1))

# Compute performance metrics
lm_dose_metrics <- lm_dose_results %>%
  metrics(truth = Y, estimate = .pred)

# Print metrics
lm_dose_metrics
```
The linear model with DOSE as the only predictor explains about 51.6% of the variance in Y (R² = 0.516), with an RMSE of 666. The coefficients show a clear positive dose-response relationship. Higher doses are associated with greater drug exposure.
### Fit a linear model for Y using all predictors
```{r}
# Fit a linear model with all predictors
lm_all_fit <- lm_spec %>%
  fit(Y ~ DOSE + AGE + SEX + RACE + WT + HT, data = d1)

# View model results
lm_all_fit

# Generate predictions from the full linear model
lm_all_results <- d1 %>%
  select(Y) %>%
  bind_cols(predict(lm_all_fit, new_data = d1))

# Compute performance metrics
lm_all_metrics <- lm_all_results %>%
  metrics(truth = Y, estimate = .pred)

# Print metrics
lm_all_metrics
```
Adding all predictors improves the model fit — R² increases to 0.62 and RMSE drops to 590. DOSE remains the strongest predictor, with a clear positive dose response relationship.
### Fit a logistic model to the SEX
```{r}
# Specify a logistic regression model
log_spec <- logistic_reg() %>%
  set_engine("glm")

# Fit a logistic model with DOSE as the predictor
log_dose_fit <- log_spec %>%
  fit(SEX ~ DOSE, data = d1)

# View model results
log_dose_fit

# Generate class predictions and probabilities, then combine with truth
log_dose_results <- d1 %>%
  select(SEX) %>%
  bind_cols(
    predict(log_dose_fit, new_data = d1, type = "class"),
    predict(log_dose_fit, new_data = d1, type = "prob")
  )

# Compute classification accuracy
log_dose_results %>%
  accuracy(truth = SEX, estimate = .pred_class)

# Compute ROC-AUC using the first level as the event
log_dose_results %>%
  roc_auc(truth = SEX, .pred_1, event_level = "first")
```

### Fit a logistic model for SEX using all predictors
```{r}
# Fit a logistic regression model with all predictors
log_all_fit <- log_spec %>%
  fit(SEX ~ DOSE + AGE + RACE + WT + HT , data = d1)

# Generate predicted classes
log_all_class <- predict(log_all_fit, new_data = d1, type = "class")

# Generate predicted probabilities
log_all_prob <- predict(log_all_fit, new_data = d1, type = "prob")

# Combine results
log_all_results <- d1 %>%
  select(SEX) %>%
  bind_cols(log_all_class) %>%
  bind_cols(log_all_prob)

# Print the fitted model
log_all_fit

# Compute accuracy
log_all_results %>%
  accuracy(truth = SEX, estimate = .pred_class)

# Compute ROC AUC
log_all_results %>%
  roc_auc(truth = SEX, .pred_1, event_level = "first")
```
The logistic model using DOSE alone showed poor predictive performance, with a ROC-AUC of 0.592 and accuracy of 0.867, where the high accuracy is largely driven by class imbalance. Adding all predictors substantially improved the model, with ROC-AUC increasing to 0.978 and accuracy reaching 0.95. However, this does not mean sex can be clinically predicted from these variables. And since gender is something we already know, it has no practical significance.