---
title: "Fitting Exercise"
---

# Analysis of the Mavoglurant pharmacokinetic data
## Loading data and initial exploration
```{r, warning=FALSE}
# Load packages
library(tidyverse)
library(tidymodels)
library(GGally)
library(here)
# Load data
data <- read_csv(here("fitting-exercise", "Mavoglurant_A2121_nmpk.csv"))

# quick cheak
glimpse(data)
```

```{r}
# Plot DV over TIME for each individual, colored by dose
ggplot(data, aes(x = TIME, y = DV, group = ID, color = factor(DOSE))) +
  geom_line(alpha = 0.4) +
  labs(
    title = "Drug concentration over time by individual",
    x = "Time",
    y = "DV",
    color = "Dose"
  )
```

```{r}
# keep one OOC = 1
data <- data %>%
  filter(OCC == 1)

# Create a new data frame with the sum of DV for each individual, excluding time 0
y_dt <- data %>%
  filter(TIME != 0) %>%
  group_by(ID) %>%
  summarize(Y = sum(DV, na.rm = TRUE), .groups = "drop")

# keep time 0 data
time0 <- data %>%
  filter(TIME == 0) 

# Merge the two data frames to include the sum of DV and the time 0 data
d1 <- time0 %>%
  left_join(y_dt, by = "ID")

# Check the data
dim(d1) 
glimpse(d1)

# Converts RACE and SEX to factor variables and keeps only these variables: Y,DOSE,AGE,SEX,RACE,WT,HT
d1 <- d1 %>%
  mutate(
    RACE = as.factor(RACE),
    SEX = as.factor(SEX)
  ) %>% 
  select(Y, DOSE, AGE, SEX, RACE, WT, HT)

# Check the data
glimpse(d1)
head(d1)
```

## Exploratory data analysis

# Conclusion
From the results, it can be seen that in both models, the risk of dementia increases with age and depression, while the risks of diabetes and hypertension increase. Adding the secondary BMI term significantly improves the model's fit and confirms the expected non-linear (U-shaped) relationship between BMI and the risk of dementia.